---
title: "STAT 243 Final Project: Adaptive Rejection Sampling"
author: "Yuqi Yang, Shuyao Wang and Nolan David Gunter"
date: "12/15/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ars)
library(testthat)
library(assertthat)
library(numDeriv)
source("../R/utility.R")
source("../R/ars.R")
```

## Github Package

 - The Github page can be accessed at: https://github.berkeley.edu/yuqiyang/ars
 - The .tar.gz file for the package can be accessed at: https://github.berkeley.edu/yuqiyang/ars/blob/main/ars_0.1.0.tar.gz


## Description

Our implementation of adaptive rejection sampling is designed to encourage 
vectorized and modularized code for efficiency. 

The `utility.R` file includes a variety of utility functions used to conduct a 
single task, most of which to called in `ars.R` where the `ars()` function 
available to users is defined. 

- `zfindj()` is a helper function that finds $z_{j-1}$ (stored as `z[j]`) and 
$z_j$ (stored as `z[j+1]`) such that $x\in[z_{j-1}, z_j]$, and returns `j`
- `u_fun()` uses the definition of the $u_k(x)$ function given by equation 
$(2)$ from Gilks and Wild.
 
Next, in order to draw random samples, we derive the inverse CDF from the 
equation $(3)$ given in Gilks and Wilk, which is:
$$
s_k(x) = \frac{\exp(u_k(x))}{\int_D \exp{u_k(x')}dx'}
$$

We can calculate the integral in the denominator analytically:
$$
\int_D \exp{u_k(x')}dx' = \sum_{j=1}^k \int_{z_{j-1}}^{z_j} \exp{u_k(x')}dx' 
$$
$$
 = \sum_{j=1}^k \int_{z_{j-1}}^{z_j} \exp(x'h'(x_j) + h(x_j) - x_jh'(x_j))dx'
$$

$$
 = \sum_{j=1}^k [\frac{1}{h'(x_j)} \exp{u_k(x')}]_{z_{j-1}}^{z_j} 
$$
$$
 = \sum_{j=1}^{k} \frac{1}{h'(x_j)} (\exp{u_k(z_j)} - \exp{u_k(z_{j-1})})
$$

Note that the denominator is constant, so we may define it as $C$. Thus
$$
s_k(x) = \frac{exp(u_k(x))}{C} = \frac{exp(xh'(x_j) + h(x_j) - x_j h'(x_j))}{C}
$$

Note that $s_k(x)$ is the pdf of the probability distribution. To find the CDF 
($Pr(X < x)$), we may take the integral $\int_{z_0}^{x} s_k(t) dt$, where $z_0$ 
is the lower bound. 

Since $s_k(x)$ is piece-wise exponential, we can calculate the integral in a 
piece-wise way:
$$
p = Pr(X < x) = \int_{z_0}^x s_k(t) dt = \sum_{i=1}^{j-1}\int_{z_{i-1}}^{z_i} s_k(t)dt + \int_{z_{j-1}}^x s_k(t)dt 
$$

$$
= \sum_{i=1}^{j-1}\int_{z_{i-1}}^{z_i} \frac1C \exp(th'(x_i)+h(x_i)-x_ih'(x_i)) dt + \int_{z_{j-1}}^x \frac1C \exp(th'(x_j)+h(x_j)-x_jh'(x_j))dt
$$

$$
= \sum_{i=1}^{j-1} \frac{1}{Ch'(x_i)}(\exp{u_k(z_i)}-\exp{u_k(z_{i-1})}) + \frac{1}{Ch'(x_j)}(\exp{u_k(x)}-\exp{u_k(z_{j-1})})
$$

In order to draw samples, we may draw $p$ from a uniform distribution of $[0,1]$ 
and use the inverse CDF to find the corresponding $x$. The inverse CDF is simply 
solving the above equation for $x$ given $p$:
$$
\frac{1}{Ch'(x_j)}(\exp{u_k(x)}-\exp{u_k(z_{j-1})})
$$

$$
= p - \sum_{i=1}^{j-1} \frac{1}{Ch'(x_i)}(\exp{u_k(z_i)}-\exp{u_k(z_{i-1})}) 
\exp{u_k(x)}-\exp{u_k(z_{j-1})} 
$$

$$
= Ch'(x_j) (p - \sum_{i=1}^{j-1} \frac{1}{Ch'(x_i)}(\exp{u_k(z_i)}-\exp{u_k(z_{i-1})})) 
$$

$$
\exp{u_k(x)} = \exp{u_k(z_{j-1})} + Ch'(x_j) (p - \sum_{i=1}^{j-1}\frac{1}{Ch'(x_i)}(\exp{u_k(z_i)}-\exp{u_k(z_{i-1})})) u_k(x) 
$$

$$
= \log(\exp{u_k(z_{j-1})} + Ch'(x_j) (p - \sum_{i=1}^{j-1} \frac{1}{Ch'(x_i)}(\exp{u_k(z_i)}-\exp{u_k(z_{i-1})})))x
$$

$$
= (u_k(x) - h(x_j) + x_jh'(x_j)) / h'(x_j)
$$

$$
\int_{z_0}^{x} s_k(t) dt = \sum_{i=1}^{j-1} \int_{z_{i-1}}^{z_i} s_k(t) dt + 
\int_{z_{j-1}}^{x} s_k(t) dt
$$

$$
 = \sum_{i=1}^{j-1} \int_{z_{i-1}}^{z_i} \frac{exp(xh'(x_i) + h(x_i) - x_ih'(x_i))}{C} dt + \int_{z_{j-1}}^{x} \frac{exp(xh'(x_j) + h(x_j) - x_j) h'(x_j)}{C} dt
$$

$$
 = \sum_{i=1}^{j-1} \frac{exp(u_k(z_i)) - exp(u_k(z_{i-1}))}{C h'(x_i)} + \frac{exp(u_k(x)) - exp(u_k(z_{j-1}))}{C h'(x_i)}
$$

Since the above process is complicated, we break it down into several functions:
- `calculateC()` calculates the denominator $C$ for $s_k(x)$ as defined above.
- `pfindj()` finds $j$ such that the cumulative probability up to $z_{j-1} <= p < $ 
cumulative probability up to $z_j$ (denoted `z[j]` and `z[j+1]` respectively in code) 
and returns a list of `j` and the cumulative probability up to $z_{j-1}$ (`z[j]`)
- `sinvcdf_fun()` takes in the probability $p$ and returns the corresponding $x$ 
such that $Pr(X < x) = p$ given the distribution of $s_k(x)$

- `xfindj()` is a helper function for `l_fun()` that finds $x_j$ and $x_{j+1}$ 
such that $x_j <= x < x_{j+1}$. If $x < x_1$ or $x > x_k$, the function returns $0$
- `l_fun()` is the squeezing function defined by equation $(4)$ in Gilks and Wild. 
 
- `isConcave()` tests if a function is concave according to the following definition, 
in steps of 0.01 size for $\lambda$:

A function $f$ is concave in an interval $[x_1,x_2]$ for all $\lambda \in [0,1]$ if:
$$
f(\lambda x_1 + (1-\lambda)x_2) - \lambda f(x_1) - (1-\lambda) f(x_2) \geq 0
$$

For example, we know that the log of a normal density should be concave, but the log of a cauchy density will not be. 
```{r}


log.norm <- function(x) log(dnorm(x))

isConcave(log.norm, 0, 10)
```

```{r}
log.cauchy <- function(x) log(dcauchy(x))

isConcave(log.cauchy, 0, 10)
```

## Function Examples

Let's consider conducting adaptive rejection sampling in the following scenarios. 
First, we sample from a truncated normal distribution from -3 to 5.

```{r}
set.seed(1)
samps1 <- ars(g = dnorm, n = 1e2, lower = -3, upper = 5)
hist(samps1)
```

In this example, we don't supply lower or upper limits and see if the function is
still able to set default bounds. 
```{r}
set.seed(1)
samps2 <- ars(g = dnorm, n = 1e2)
hist(samps2)
```

In this example, we sample from a truncated exponential distribution from 0 to 10.
```{r}
set.seed(1)
samps3 <- ars(g = dexp, n = 1e2, lower = 0, upper = 10)
hist(samps3)
```


## Contributions

All group members read the paper and wrote their own pseudo-code to try to 
understand the adaptive rejection sampling algorithm, and met multiple times over 
zoom. 

- Nolan: wrote the initial utility functions in utility.R; troubleshooted errors 
in utility.R and ars.R; wrote descriptions and examples portions of report, and ars.Rd.
- Shuyao: wrote the initial test.R for test cases in test folder; troubleshooted 
errors in ars.R, utility.R, and report; and help finalize and connect different parts.
- Yuqi: wrote the initial main functions in ars.R; troubleshooted errors in utility 
functions and testing cases; and helped the group understand the piece wise.


## Sources

Gilks, W.R. and Wild, P. Adaptive Rejection Sampling for Gibbs Sampling. 
*Journal of the Royal Statistical Society*. 41,2, 337:248. 1992. 
